---
title: "Generating Person-Count and Person-Time Tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating Person-Count and Person-Time Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
Sys.setenv(OMP_THREAD_LIMIT = 1) # Reducing core use, to avoid accidental use of too many cores
library(Colossus)
library(data.table)
library(parallel)
```

## General Purpose

There are multiple reasons why individual level data may be grouped. It could be for visualizing the data. It can be very informative to break columns into categories and summarize the number of events or average column values in each category. It may also be to perform analysis on aggregates of the data, to reduce noise or remove identifying health information. Either way the process is similar, we want to take a table and apply categorizations and summarize events. `Colossus` offers two functions designed to create grouped tables, [Event_Count_Gen()](../reference/Event_Count_Gen.html) and [Event_Time_Gen()](../reference/Event_Time_Gen.html) which create person-count tables and person-time tables respectively. The following sections will cover generally how to use each function and what different needs are met.

### Person-Count Tables

A person count table is the simplest way to summarize a table. This assumes you have columns you want to break into categories and columns you want to summarize. Let us start with a basic example. Suppose we have a dataset tracking the number of apples and oranges people bought, the number of hands they used to carry their bag, and whether the bag ripped or not.

```{r, eval=TRUE}
apples <- c(0, 1, 2, 3, 4, 5, 6, 2, 2, 3, 4, 2, 1, 5, 6, 4, 2)
oranges <- c(1, 2, 3, 4, 5, 6, 7, 6, 5, 4, 3, 2, 1, 3, 2, 2, 1)
rip <- c(0, 1, 0, 1, 0, 1, 1, 0, 1, 1, 1, 0, 0, 0, 1, 0, 1)
hands <- c(1, 1, 2, 3, 2, 1, 2, 2, 2, 2, 1, 1, 1, 1, 1, 2, 2)
table <- data.table::data.table(
  apples = apples,
  oranges = oranges,
  rip = rip,
  hands = hands
)
```

Suppose we want to split our dataset up by the number of each fruit, count the number of ripped bags in each category, and figure out the average number of hands used. We would start by defining how we want to split the categories. There are two general ways to define the intervals, with a single string or a list of upper and lower bounds.

Let us start with the string representation. The string is expected to alternate number and delimiter, either "/" or "]", which split the upper and lower bounds for each interval. Every interval in this format includes the lower bound. The interval includes the upper bound if "]" is used, and does not otherwise. So the interval [2,5) is equivalent to "2 / 5" and the interval [5,8] is equivalent to "5 ] 8". This method uses each triplet sequentially, so the categories will cover every value that falls between the first and last number listed. The functions assume that the bounds are strictly non-decreasing, so the upper boundary is equal to the lower boundary or greater. When the upper and lower boundary are equal, the category is upper limit inclusive. In general, data are assigned to the first category they belong to, so points on an included upper boundary will be assigned to the lower category.

The list notation is very similar. There are 2 values required in the list and 1 optional value. The list has to have vectors titled lower and upper, which are the lower and upper bounds for each interval. This method can include gaps. Similarly to the string version, the lower bound is always included and the upper bound is included if the upper bound entry includes "]". The list option allows the categories to be labeled instead of being automatically numbered. A third list titled name can be provided to label each category. The string representation can name categories by including the name after the delimiter. For example, a category containing [3,5) named "middle" could be represented "3 / middle 5". Note that the name and boundary must be split by a space, and the name cannot include "/" or "]".

Suppose we want to break the apple column into the intervals: [0, 3), [3, 5), [5,7] and break our oranges column into the intervals: [-1,3), [3,6), [6,10) with the labels: few, good, and excessive. That would look like the following code. This splits each category into 3 levels and in total 9 combinations.

```{r, eval=TRUE}
# We can make our apple categories change at 0, 3, 5, 7
# We can also make category 3 include the upper value
apple_category <- "0/3/5]7"

# We can also use a list of categories. The list should have entries
# for the lower bounds, upper bounds, and optionally the names for
# each category level
orange_category <- list(
  lower = c(-1, 3, 6),
  upper = c(3, 6, 10),
  name = c("few", "good", "excessive")
)

# Finally we combine them into a list. The list names should be column
# names that the categories will be applied to.
categ <- list(
  apples = apple_category,
  oranges = orange_category
)
# We could also rename a category using `column AS new_name` in the list names.
# This is demonstrated later in the vignette.
```

Now we define what summary variables we want. Once again, we define a list. Every item in the list is named with the column it is applied to and contains a string specifying what summary method to use. There are many options supported currently, two common for a person-event table are count and mean, which will provide the sum or the mean of the column in each category. One can also provide a new name for the summary column by listing the method as "option AS name". Otherwise, the grouped table will use the original column name for the summary column name. Suppose we wanted to take the sum of ripped bags and name it "dropped" and take the mean of the hands column.

```{r, eval=TRUE}
# We might want two events to be summarized, the 'rip' and 'hands' columns.
# The first we can take the sum of, and name the summary 'dropped'
# The second we can just take a mean of
# This creates the event list, once again with list names
# matching columns in the data.
event <- list(rip = "count AS dropped", hands = "mean")
```

The final step is to run the function. The function returns two items, the grouped data and a summary of the category intervals. The category intervals should be checked to verify that the intervals match what were expected.

```{r, eval=TRUE}
# We take our table, and apply the categorizations and event summaries
# The result has two values: 'df' and 'bounds'
# df is the final grouped table, with the categories and summaries
# bounds is a description of the category levels applied
Event_Count_Gen(table, categ, event)
```

### Person-Time Tables

The previous method treats every row as being weighted the same. In many cases, every row has some measure of time which we may want to use. That is the fundamental difference between the person count tables and person time tables, the inclusion of time. The code used to generate a person time table is very similar to the person count table, with several important differences. The first is that the data should have time columns. Similar to the Cox model options in `Colossus`, the data should have an entry column, exit column, or both. These would denote the scenario where every interval ends at the same point, starts at the same point, or has different start and stop times. There are two formats of time allowed, calendar and user scale.

Calendar time data is listed in day, month, and year columns which are combined into dates. User time data is interpreted as durations observed, without any inherent units. These apply to both a person-year list, which specifies the entry and exit columns, as well as the time categories which split the data into intervals. For the calendar time scale, dates will default to 1/1/1900 if not provided. The person-year list for a calendar time-scale can also contain a unit, defaulted to years. For the following example, we will use a calendar time scale.

#### Calendar Time Scale

```{r, eval=TRUE}
# We can start by making some fake data
# a and b are arbitrary factors
# c might be the event status we are interested in
a <- c(0, 1, 2, 3, 4, 5, 6, 2, 2, 3, 4, 2, 1, 5, 6, 4, 2)
b <- c(1, 2, 3, 4, 5, 6, 7, 6, 5, 4, 3, 2, 1, 3, 2, 2, 1)
c <- c(0, 1, 0, 1, 0, 1, 1, 0, 1, 1, 1, 0, 0, 0, 1, 0, 1)

# Next we define columns for the calendar dates of each interval
# In this case we are defineing day, month, and year
d <- c(1, 1, 2, 2, 1, 1, 2, 2, 3, 3, 3, 4, 4, 2, 1, 1, 2)
e <- c(1, 1, 1, 2, 2, 2, 2, 1, 1, 1, 1, 2, 2, 2, 2, 1, 1)
f <- c(
  1900, 1900, 1900, 1900, 1900, 1900, 1900, 1900, 1900,
  1900, 1900, 1900, 1900, 1900, 1900, 1900, 1900
)
g <- c(4, 4, 4, 5, 5, 5, 5, 6, 6, 6, 5, 5, 5, 4, 4, 4, 4)
h <- c(6, 4, 4, 6, 6, 6, 4, 4, 4, 6, 6, 6, 6, 4, 4, 4, 4)
i <- c(
  1901, 1902, 1903, 1904, 1905, 1906, 1907, 1903, 1904,
  1903, 1904, 1910, 1903, 1904, 1903, 1904, 1910
)

# Finally we have our table that we want to group.
table <- data.table::data.table(
  a = a, b = b, c = c,
  d = d, e = e, f = f,
  g = g, h = h, i = i
)

# Our person-years are described by calendar dates
# We have entry and exit lists, each with the columns that
# correspond to the day/month/year.
# We also include a unit, to say we want our final person-years
# in terms of years, instead of days or months.
pyr <- list(
  entry = list(year = "f", month = "e", day = "d"),
  exit = list(year = "i", month = "h", day = "g"),
  unit = "years"
)
```

The second major difference is the possible use of time categories. For a calendar time scale, time categories are either set by calendar date or by age. Calendar time scales are set by vectors of the day, month, and year of each interval splitting point. So intervals changing every 2 days could be expressed by "day=c(1,3,5,...)". Age time scales use a day/month/year as a reference for age, and then set the age categories using the same notation as non-time categories (either a single string or list of upper and lower bounds). Unlike the non-time categories, we can not name the categories and any attempt to do will be ignored and produce a warning. When age categories are used, a new column is created for the average `age` in each interval.

Time categories are unique in that each row can correspond to multiple time categories, so the duration of each row in each time category is kept to distinguish which rows are only partially in each time interval. If a time category is not provided, then the full duration of each row is used for person-year calculations. Multiple categories can be used for a calendar time scale, however it should be noted that multiple calendar date based categories could be combined into a single list of categories.

```{r, eval=TRUE}
# The non-time categories are defined as previously shown.
categ <- list(
  a = "-1/3/5]7",
  b = list(
    lower = c(-1, 3, 6), upper = c(3, 6, 10),
    name = c("low", "medium", "high")
  )
)

# The time categories follow a similar structure.
# We have a list with names that correspond to columns,
# and entries that give category limits.
# Our calendar categories are defined between each of these dates,
# upper bound exclusive.

# Time categories make a new column for the category, which should not be in the data.
time_scale <- list(`time_bin` = list(
  day = c(1, 1, 1),
  month = c(1, 1, 1),
  year = c(1899, 1903, 1910)
))
```

The final difference is that person-time tables make a distinction between events and summaries. The summary list is nearly identical to the event list in the person-count function, with the exception of allowing for means to be weighted by person-years. The event list for the person-time table function is used to determine which time categories the event columns are assigned to. By default, the number of events in each grouping is recorded in the final table. Based on the relative row intervals and category intervals, each row can be part of multiple time categories. However, the event is only assigned to the time category containing the endpoint of the row interval. Person-time tables also return the total intervals "at risk" in each category. Observations that are split into multiple time categories will contribute to the number at risk.

```{r, eval=TRUE}
# Summaries follow the person-count table notation for events.
# We apply some summarization to an existing column.
# The name should be the column, and we can use 'AS'
# to rename the summary as needed.
summary <- list(b = "weighted_mean AS b_weighted")

# Events in a person-time table are assumed to occur at the end
# of the interval.
# Events are automatically added to the summary list as a sum.
events <- list("c AS cases")
```

Finally, the function is run. Similar to the person-count table function, the function returns the grouped table and list of category boundaries.

```{r, eval=TRUE}
# We run the analysis just like the event_count data
# Here we are also setting fcount to true, to return the first at risk in each group
# We get the same returned value, the table and summary of category boundaries
print(Event_Time_Gen(table, pyr, time_scale, categ, summary, events, fcount = T))

# We can also change the units in our person-years
pyr <- list(
  entry = list(year = "f", month = "e", day = "d"),
  exit = list(year = "i", month = "h", day = "g"),
  unit = "days"
)
print(Event_Time_Gen(table, pyr, time_scale, categ, summary, events, fcount = T))
```

#### User Defined Time Scale

As previously mentioned, the time scale can also be durations with arbitrary units. This is particularly important if the study is using follow-up time and there is no need to translate to calendar points. The differences start with the person-year list. In this case the entry and exit items are just columns, the duration at entry and exit of the observation.

```{r, eval=TRUE}
# Repeating the same non-time categories and event statuses
a <- c(0, 1, 2, 3, 4, 5, 6, 2, 2, 3, 4, 2, 1, 5, 6, 4, 2)
b <- c(1, 2, 3, 4, 5, 6, 7, 6, 5, 4, 3, 2, 1, 3, 2, 2, 1)
c <- c(0, 1, 0, 1, 0, 1, 1, 0, 1, 1, 1, 0, 0, 0, 1, 0, 1)

# Instead of calendar dates, we may have entry and exit durations
d <- c(1, 1, 2, 2, 10, 10, 15, 15, 20, 20, 30, 30, 40, 40, 50, 50, 60)
e <- c(11, 11, 22, 22, 30, 30, 45, 45, 50, 50, 60, 60, 70, 70, 80, 80, 90)

table <- data.table::data.table(
  a = a, b = b, c = c,
  d = d, e = e
)

# When we have durations instead of calendar dates, our person-year
# list should only have a single entry, instead of lists of columns.
pyr <- list(entry = "d", exit = "e")
```

Next, we define the time categories. The only accepted time category for a user defined scale is similar to the age-based calendar scale categories. It is assumed to be provided as a single string or a list with upper and lower limits. While we can technically provide multiple categories, they can always be condensed into a single category similar to the calendar categories.

```{r, eval=TRUE}
# Once again, our non-time categories are defined the same
categ <- list(
  "a" = "-1/-1/3/5]7",
  "b AS b_bin" = list(lower = c(-1, -1, 3, 6), upper = c(-1, 3, 6, "]10"))
)

# Here we are defining the duration categories to use.
# This will create two columns, 'time' and 'time_bin'.
# 'time' is the average duration between the entry and exit durations.
# 'time_bin' is the category value.
# We can define a representative duration to find the average duration in each group.
time_scale <- list("time AS time_bin" = "0 / 10 / 50 / 100")
```

Finally, there are functionally no differences between the summaries and events available for person-time tables using calendar and user defined time scales.

```{r, eval=TRUE}
summary <- list("c" = "count AS cases", "a" = "mean", "b" = "weighted_mean")
events <- list("c")

# Everything else runs the same, and gives similar outputs.
# Except now the time category boundaries are not based on dates.
print(Event_Time_Gen(table, pyr, time_scale, categ, summary, events, T))
```

### Additional summary options

There are multiple options for optional returned summaries available for both person-count and person-time tables. Person-time tables calculate the typical "at risk" summary variables for any row at risk, but they also can calculate "first at risk" and "last at risk" summary variables (named `F_AT_RISK` and `L_AT_RISK` respectively). These values can be summarized by setting their corresponding function parameters, `fcount` and `lcount` to true. Note that these are not calculated during run-time otherwise, so they can be used for taking weighted means only if they are returned. Notably, these summaries determine the first and last observed time category for each subject. This means that even if a subject is observed past the last time category, they are still considered "last at risk" during the final time category. Similarly for the first at risk measure. These summaries will always sum to the total subjects at risk at any point during the time categories. These summaries can also be extended to subject IDs, if there are multiple intervals for each subject. By passing a `studyid` parameter to the person-time function, the at risk summaries will summarize per ID, instead of per observed interval.

In addition to the summary and event methods listed in this vignette, there are 11 methods allowed by person-time tables and 7 methods allowed by person-count tables. As mentioned previously, methods can be written as "option AS name" to rename the output, but they can also be weighted by using "option BY covariate" (or both renamed and weighted by using both). Weighting can be applied to turn almost every non-weighted summary into a weighted version. Weighted summaries can be listed without a weighting column for person-time tables to default to weighting by person-years, however there is no default for person-event tables.

| Option | Description | Alternative Writing | Available for Person-Count |
| :---------: | :----------: | :----------: | :----------: |
| count, sum, rsum | Takes the sum | N/A | Yes |
| mean, rmean | Takes the mean | N/A | Yes |
| weighted_mean | Takes the weighted mean, either by person-years or by a column selected using BY | `mean BY column` or `rmean BY column` | Yes |
| weighted_sum | Takes the weighted sum, either by person-years or by a column selected using BY | `sum BY column`, `rsum BY column`, or `count BY column` | Yes |
| xsum | Takes the sum over the last at risk intervals | N/A | No |
| xmean | Takes the mean over the last at risk intervals | N/A | No |
| weighted_xsum | Takes the weighted sum over the last at risk intervals, either by person-years or by a column selected using BY | `xsum BY column` | No |
| weighted_xmean | Takes the weighted mean over the last at risk intervals, either by person-years or by a column selected using BY | `xmean BY column` | No |




